quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-18 paper-quollr.Rmd
Developed the outputs with detour (links)

1. **Data preparation** — functions such as `scale_highd()` and `prepare_data()` standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
2. **Hexagonal binning** — the `hex_bin_data()` module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
3. **Model fitting** — the `fit_highd_model()` function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
4. **Error computation** — functions like `compute_errors()` and `aggregate_bin_error()` evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
5. **Visualization and linkage** — plotting and tour functions (`show_hex_model()`, `show_link_plots()`, and `show_error_link_plots()`) integrate outputs into linked static and dynamic displays using `ggplot2`, `langevitour`, and `crosstalk`.

Each module is internally independent but communicates through standardized data objects (see next section). This modular design simplifies maintenance and allows developers to extend individual components—such as substituting different binning strategies, embedding methods, or visualization tools—without altering the overall workflow.

@@

detour(
  df_exe,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  show_scatter(axes = TRUE, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"))

@@

As an alternative to generate tour, the link plots can be generate with `detourr` as well. To achieve this, users can construct the linked visualization manually using the `crosstalk` and `htmltools` packages. The two- and three-panel linked layouts are created by arranging the NLDR plot (`nldr_plt`), the optional error distribution plot (`error_plt`), and a dynamic high-dimensional projection view (`langevitour_output` or `detourr_output`) side by side within a flexible grid layout. The `bscols()` function from `crosstalk` synchronizes point selection between these plots, ensuring linked brushing across visual components.

@@

This article is created using \CRANpkg{knitr} [@yihui2015] and \CRANpkg{rmarkdown} [@yihui2018] in R with the `rjtools::rjournal_article` template. These `R` packages were used for this work: `cli` [@gabor2025], `dplyr` [@hadley2023], `ggplot2` [@hadley2016], `interp` (>= 1.1-6) [@albrecht2024], `langevitour` [@paul2023], `proxy`[@david2022], `stats` [@core2025], `tibble` [@kirill2023], `tidyselect` [@lionel2024], `crosstalk` [@joe2023], `plotly` [@chapman2020], `kableExtra` [@hao2024], `patchwork` [@thomas2024], and `readr` [@hadley2024].

>>>>

1. Data preparation — functions such as `gen_scaled_data()` standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
2. Hexagonal binning — the `hex_bin_data()` module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
3. Model fitting — the `fit_highd_model()` function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
4. Error computation — functions like `compute_errors()` and `aggregate_bin_error()` evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
5. Visualization and linkage — plotting and tour functions (`show_hex_model()`, `show_link_plots()`, and `show_error_link_plots()`) integrate outputs into linked static and dynamic displays using `ggplot2`, `langevitour`, and `crosstalk`.

Each module is internally independent but communicates through standardized data objects (see next section). This modular design simplifies maintenance and allows developers to extend individual components such as substituting different binning strategies, embedding methods, or visualization tools without altering the overall workflow.

@@

detour(
  df_exe,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  tour_path(grand_tour(2), 
                    max_bases=50, fps = 60) |>
  show_scatter(axes = TRUE, size = 0.5, alpha = 0.5, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"),
               width = "600px", height = "600px")
```

@@

As an alternative to using `langevitour`, link plots can also be generated with the `detourr` package. In this setup, users can manually construct the linked visualization using the `crosstalk` [@joe2023] and `htmltools` [@joe2024] packages. The interactive layout is created by arranging the \twoD NLDR plot (`nldr_plt`), the optional error distribution plot (`error_plt`), and the tour view produced with `detourr`, side by side within a flexible grid. The NLDR and error plots are rendered with `ggplotly()` [@chapman2020] to enable interactive linking. The `bscols()` function from `crosstalk` manages synchronization across these panels, allowing for linked brushing and coordinated selection between these interactive plots.

A two-panel linked plot combining the NLDR view and the tour from `detourr` can be created as follows:

```{r}
point_colours <- c("#66B2CC", "#FF7755")
point_sizes <- c(0, 1)

num_highd_col <- df_exe |>
  dplyr::select(starts_with("x")) |>
  NCOL()

df_all <- df_exe |>
  dplyr::filter(type == "data") ## original dataset

shared_df <- crosstalk::SharedData$new(df_exe)

nldr_plt <- shared_df |>
  ggplot(aes(x = emb1, y = emb2)) +
  geom_point(alpha=0.5, colour=point_colours[1], size = 0.5) +
  theme_linedraw() +
  theme(
    #aspect.ratio = 1,
    plot.background = element_rect(fill = 'transparent', colour = NA),
    plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
    panel.background = element_rect(fill = 'transparent',
                                    colour = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.y = element_blank()
  )

nldr_plt <- ggplotly(nldr_plt, width = 600,
                     height = 600, tooltip = "none") |>
  style(unselected=list(marker=list(opacity=1))) |>
  highlight(on="plotly_selected", off="plotly_deselect") |>
  config(displayModeBar = FALSE)

```{r, echo=TRUE, eval=knitr::is_html_output(), layout = "l-page"}

detourr_output <- detour(
  shared_df,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  tour_path(grand_tour(2), 
                    max_bases=50, fps = 60) |>
  show_scatter(axes = TRUE, size = 0.5, alpha = 0.5, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"),
                width = "600px", height = "600px")

crosstalk::bscols(
    htmltools::div(style="display: grid; grid-template-columns: 1fr 1fr;",
                   nldr_plt,
                   htmltools::div(style = "margin-top: 20px;", detourr_output)
    ),
    device = "xs"
  )
```

For analyses that include model error visualization, a three-panel view can be constructed by adding the error distribution plot (`error_plt`):

```{r}

num_highd_col <- df_exe |>
  dplyr::select(starts_with("x")) |>
  NCOL()

df_all <- df_exe |>
  dplyr::filter(type == "data") ## original dataset

shared_df <- crosstalk::SharedData$new(df_exe)

error_plt <- shared_df |>
  ggplot(aes(x=sqrt_row_wise_total_error, y = density)) +
  geom_point(colour = point_colours[1]) +
  xlab(expression(e[hj])) +
  ylab("") +
  theme_bw() +
  theme(
    #aspect.ratio = 1,
    plot.background = element_rect(fill = 'transparent', colour = NA),
    plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
    panel.background = element_rect(fill = 'transparent',
                                    colour = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.y = element_blank()
  )

error_plt <- ggplotly(error_plt, width = 400,
                      height = 400, tooltip = "none") |>
  style(unselected=list(marker=list(opacity=1))) |>
  highlight(on="plotly_selected", off="plotly_deselect") |>
  config(displayModeBar = FALSE)

nldr_plt <- shared_df |>
  ggplot(aes(x = emb1, y = emb2)) +
  geom_point(alpha=0.5, colour=point_colours[1], size = 0.5) +
  theme_linedraw() +
  theme(
    #aspect.ratio = 1,
    plot.background = element_rect(fill = 'transparent', colour = NA),
    plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
    panel.background = element_rect(fill = 'transparent',
                                    colour = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.y = element_blank()
  )

nldr_plt <- ggplotly(nldr_plt, width = 400,
                     height = 400, tooltip = "none") |>
  style(unselected=list(marker=list(opacity=1))) |>
  highlight(on="plotly_selected", off="plotly_deselect") |>
  config(displayModeBar = FALSE)
```

```{r, echo=TRUE, eval=knitr::is_html_output(), layout = "l-page"}

detourr_output <- detour(
  shared_df,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  tour_path(grand_tour(2), 
                    max_bases=50, fps = 60) |>
  show_scatter(axes = TRUE, size = 0.5, alpha = 0.5, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"),
                width = "600px", height = "600px")

crosstalk::bscols(
  htmltools::div(
    style = "display: grid; grid-template-columns: 1fr 1fr 1fr;",
    error_plt, 
    nldr_plt,
    htmltools::div(style = "margin-top: 20px;", detourr_output)
  ),
  device = "xs"
)
```

@@

This article is created using \CRANpkg{knitr} [@yihui2015] and \CRANpkg{rmarkdown} [@yihui2018] in R with the `rjtools::rjournal_article` template. These `R` packages were used for this work: `cli` [@gabor2025], `dplyr` [@hadley2023], `ggplot2` [@hadley2016], `interp` (>= 1.1-6) [@albrecht2024], `langevitour` [@paul2023], `proxy`[@david2022], `stats` [@core2025], `tibble` [@kirill2023], `tidyselect` [@lionel2024], `crosstalk` [@joe2023], `plotly` [@chapman2020], `htmltools` [@joe2024], `kableExtra` [@hao2024], `patchwork` [@thomas2024], and `readr` [@hadley2024].

### 2025-10-18 paper-quollr.tex
Developed the outputs with detour (links)

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Data preparation} --- functions such as \texttt{scale\_highd()} and \texttt{prepare\_data()} standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  \textbf{Hexagonal binning} --- the \texttt{hex\_bin\_data()} module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
\item
  \textbf{Model fitting} --- the \texttt{fit\_highd\_model()} function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
\item
  \textbf{Error computation} --- functions like \texttt{compute\_errors()} and \texttt{aggregate\_bin\_error()} evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
\item
  \textbf{Visualization and linkage} --- plotting and tour functions (\texttt{show\_hex\_model()}, \texttt{show\_link\_plots()}, and \texttt{show\_error\_link\_plots()}) integrate outputs into linked static and dynamic displays using \texttt{ggplot2}, \texttt{langevitour}, and \texttt{crosstalk}.
\end{enumerate}

Each module is internally independent but communicates through standardized data objects (see next section). This modular design simplifies maintenance and allows developers to extend individual components---such as substituting different binning strategies, embedding methods, or visualization tools---without altering the overall workflow.

@@

detour(
  df_exe,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  show_scatter(axes = TRUE, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"))
\end{verbatim}

@@

As an alternative to generate tour, the link plots can be generate with \texttt{detourr} as well. To achieve this, users can construct the linked visualization manually using the \texttt{crosstalk} and \texttt{htmltools} packages. The two- and three-panel linked layouts are created by arranging the NLDR plot (\texttt{nldr\_plt}), the optional error distribution plot (\texttt{error\_plt}), and a dynamic high-dimensional projection view (\texttt{langevitour\_output} or \texttt{detourr\_output}) side by side within a flexible grid layout. The \texttt{bscols()} function from \texttt{crosstalk} synchronizes point selection between these plots, ensuring linked brushing across visual components.

@@

This article is created using \CRANpkg{knitr} \citep{yihui2015} and \CRANpkg{rmarkdown} \citep{yihui2018} in R with the \texttt{rjtools::rjournal\_article} template. These \texttt{R} packages were used for this work: \texttt{cli} \citep{gabor2025}, \texttt{dplyr} \citep{hadley2023}, \texttt{ggplot2} \citep{hadley2016}, \texttt{interp} (\textgreater= 1.1-6) \citep{albrecht2024}, \texttt{langevitour} \citep{paul2023}, \texttt{proxy}\citep{david2022}, \texttt{stats} \citep{core2025}, \texttt{tibble} \citep{kirill2023}, \texttt{tidyselect} \citep{lionel2024}, \texttt{crosstalk} \citep{joe2023}, \texttt{plotly} \citep{chapman2020}, \texttt{kableExtra} \citep{hao2024}, \texttt{patchwork} \citep{thomas2024}, and \texttt{readr} \citep{hadley2024}.

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Data preparation --- functions such as \texttt{gen\_scaled\_data()} standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  Hexagonal binning --- the \texttt{hex\_bin\_data()} module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
\item
  Model fitting --- the \texttt{fit\_highd\_model()} function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
\item
  Error computation --- functions like \texttt{compute\_errors()} and \texttt{aggregate\_bin\_error()} evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
\item
  Visualization and linkage --- plotting and tour functions (\texttt{show\_hex\_model()}, \texttt{show\_link\_plots()}, and \texttt{show\_error\_link\_plots()}) integrate outputs into linked static and dynamic displays using \texttt{ggplot2}, \texttt{langevitour}, and \texttt{crosstalk}.
\end{enumerate}

Each module is internally independent but communicates through standardized data objects (see next section). This modular design simplifies maintenance and allows developers to extend individual components such as substituting different binning strategies, embedding methods, or visualization tools without altering the overall workflow.

@@

detour(
  df_exe,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  tour_path(grand_tour(2), 
                    max_bases=50, fps = 60) |>
  show_scatter(axes = TRUE, size = 0.5, alpha = 0.5, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"),
               width = "600px", height = "600px")
\end{verbatim}

@@

As an alternative to using \texttt{langevitour}, link plots can also be generated with the \texttt{detourr} package. In this setup, users can manually construct the linked visualization using the \texttt{crosstalk} \citep{joe2023} and \texttt{htmltools} \citep{joe2024} packages. The interactive layout is created by arranging the $2\text{-}D$ NLDR plot (\texttt{nldr\_plt}), the optional error distribution plot (\texttt{error\_plt}), and the tour view produced with \texttt{detourr}, side by side within a flexible grid. The NLDR and error plots are rendered with \texttt{ggplotly()} \citep{chapman2020} to enable interactive linking. The \texttt{bscols()} function from \texttt{crosstalk} manages synchronization across these panels, allowing for linked brushing and coordinated selection between these interactive plots.

A two-panel linked plot combining the NLDR view and the tour from \texttt{detourr} can be created as follows:

\begin{verbatim}
detourr_output <- detour(
  shared_df,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  tour_path(grand_tour(2), 
                    max_bases=50, fps = 60) |>
  show_scatter(axes = TRUE, size = 0.5, alpha = 0.5, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"),
                width = "600px", height = "600px")

crosstalk::bscols(
    htmltools::div(style="display: grid; grid-template-columns: 1fr 1fr;",
                   nldr_plt,
                   htmltools::div(style = "margin-top: 20px;", detourr_output)
    ),
    device = "xs"
  )
\end{verbatim}

For analyses that include model error visualization, a three-panel view can be constructed by adding the error distribution plot (\texttt{error\_plt}):

\begin{verbatim}
detourr_output <- detour(
  shared_df,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  tour_path(grand_tour(2), 
                    max_bases=50, fps = 60) |>
  show_scatter(axes = TRUE, size = 0.5, alpha = 0.5, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"),
                width = "600px", height = "600px")

crosstalk::bscols(
  htmltools::div(
    style = "display: grid; grid-template-columns: 1fr 1fr 1fr;",
    error_plt, 
    nldr_plt,
    htmltools::div(style = "margin-top: 20px;", detourr_output)
  ),
  device = "xs"
)
\end{verbatim}

@@

This article is created using \CRANpkg{knitr} \citep{yihui2015} and \CRANpkg{rmarkdown} \citep{yihui2018} in R with the \texttt{rjtools::rjournal\_article} template. These \texttt{R} packages were used for this work: \texttt{cli} \citep{gabor2025}, \texttt{dplyr} \citep{hadley2023}, \texttt{ggplot2} \citep{hadley2016}, \texttt{interp} (\textgreater= 1.1-6) \citep{albrecht2024}, \texttt{langevitour} \citep{paul2023}, \texttt{proxy}\citep{david2022}, \texttt{stats} \citep{core2025}, \texttt{tibble} \citep{kirill2023}, \texttt{tidyselect} \citep{lionel2024}, \texttt{crosstalk} \citep{joe2023}, \texttt{plotly} \citep{chapman2020}, \texttt{htmltools} \citep{joe2024}, \texttt{kableExtra} \citep{hao2024}, \texttt{patchwork} \citep{thomas2024}, and \texttt{readr} \citep{hadley2024}.
